### 2021-07-29 paper/paper.tex
updated results and added paper

 \documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
%\usepackage{hyphenat}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage[binary-units=true]{siunitx}
\usepackage{ulem}
%\usepackage{censor}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{tabularx}

\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{The benefits of prefetching for large-scale cloud-based neuroimaging analysis workflows}

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan: }#1\color{black}}
\newcommand{\tristanmod}[2]{\color{orange}\sout{#1}{#2}\color{black}}

\newcommand{\ariel}[1]{\color{blue}\textbf{From Ariel:}#1\color{black}}
\newcommand{\arielmod}[2]{\color{blue}\sout{#1}{#2}\color{black}}

\newcommand{\valerie}[1]{\color{purple}\textbf{Valerie: }#1\color{black}}
\newcommand{\valeriemod}[2]{\color{purple}\sout{#1}{#2}\color{black}}

\newcommand{\aws}{AWS\xspace}
\newcommand{\hcp}{HCP\xspace}
\newcommand{\sfs}{S3Fs\xspace}

\author{Valerie Hayot-Sasson$^1$, Tristan Glatard$^1$, Ariel Rokem$^2$ \\ $^1$ Department of Computer-Science and Software Engineering, Concordia University, Montreal, Canada\\ $^2$ Department of Psychology and eScience Institute, University of Washington, Seattle, Washington, USA}

\maketitle
\begin{abstract}
%%edited
To support the growing demands of neuroscience applications, researchers are transitioning to cloud computing for its scalable, robust and elastic infrastructure. Nevertheless, large datasets residing in object stores may result in significant data transfer overheads during workflow execution.
Prefetching, a method to mitigate the cost of reading in mixed workloads, masks data transfer costs within processing time of prior tasks. We present an implementation of ``Rolling Prefetch", a Python library that implements a particular form of prefetching from AWS S3 object store, and we quantify its benefits.

Rolling Prefetch extends \sfs, a Python library exposing AWS S3 functionality via a file object, to add prefetch capabilities. 
In measured analysis performance of a 500~GB brain connectivity dataset stored on S3, we found that prefetching provides significant speed-ups of up to 1.8$\times$, even in applications consisting entirely of data loading. The observed speed-up values are consistent with our theoretical analysis. Our results demonstrate the usefulness of prefetching for scientific data processing on cloud infrastructures and provide an implementation applicable to various application domains.

%%%%%%%%%%%%%%%% original

    % With the increase in dataset size and computing requirements of neuroimaging workflows, researchers have transitioned to cloud environments to provide the necessary infrastructure to enable the research. Since compute-local storage on the cloud can be rather costly,
    % large datasets are placed on slower, but scalable,
    % cloud storage instances, resulting in non-negligible performance penalties when accessing the data during workflow execution.
    % A popular method for mitigating the cost of data accesses is through prefetching. In prefetching, the data is transferred fast storage in
    % anticipation of a task access, masking slow storage latency and bandwidth within task scheduling and compute time.

    % In this paper, we investigate the benefit of prefetching on a neuroimaging use case running on a t2-xlarge Amazon EC2 instance operating
    % on an 500~GB tractography dataset stored on Amazon S3. To achieve this, we designed a model to describe the anticipated performance trends, developed a Python library based on S3Fs to prefetch the data and compared its performance with S3Fs as its baseline. Overall, it was found
    % that prefetching the data provided a significant speedup when large amounts of data needed to processed by a task and when there were many
    % parallel tasks accessing the dataset.

 %%%%%%%%%%%%%%%%%%%%%
  %Neuroimaging deals with massive amounts of data. One way to deal with that is to move to cloud computing. One of the persistent bottlenecks in %data analysis in these environments is that you have to wait for data to arrive at the CPU to be able to start analyzing it. Here, we devise a %way to prefetch data that will be needed in the future, while other work is happening.

\end{abstract}

\begin{IEEEkeywords}
Prefetching, neuroimaging, cloud computing
\end{IEEEkeywords}

\section{Introduction}

% Cloud service providers, such as Amazon Web Services (\aws)~\cite{aws},
% Microsoft Azure~\cite{azure} and Google Cloud~\cite{gc}, give access to two
% main services: Compute services and Object Stores. Compute services
% typically consist of instances with a configured amount of RAM, vCPU and
% local storage.  Costs associated with compute services increase with the
% number of resources and time required. Compared to object stores,
% compute-local storage can be very costly, particularly when storing large
% amounts of data for extended periods. Thus, users generally opt for
% scalable external object stores for their large data. This is despite the
% increased latency, which may vary depending on location of the storage and
% compute instance network. Examples of cloud object storage include Amazon's
% Simple Storage Service (S3), Azure's Blob Service, and Google Cloud
% Storage.

Many fields of science are experiencing increases in the volume of datasets available to researchers. Neuroscience in particular is experiencing a rapid growth in data due to technical advances, 
scientific breakthroughs, and sociological trends towards open data sharing. This increase in data is providing the basis for new discoveries about brain structure and function, but it also presents technical challenges. To deal with the deluge of available data, neuroscientists are increasingly adopting cloud platforms for data storage and processing. However, inefficient use of cloud services can lead
to needlessly longer processing time and cost. We aim to investigate
methods to reduce the impact of data transfers for
neuroimaging workflows on cloud services.

Data prefetching
is a well-established technique for the reduction of data access-related costs~\cite{callahan1991software, mowry1992design,
klaiber1991architecture}. Traditionally, prefetching was
used to reduce memory latency, as memory accesses were significantly slower than CPU processing. However, since the rise of Big Data,
prefetching has also been shown to be beneficial to the processing of large
datasets located on remote storage~\cite{yildiz2018improving}. During the
execution of an application, data required for future tasks are
copied from the remote storage device to compute-local storage, such that
when the application requires the data, it can read it from local
storage.

A recent example of the effectiveness of prefetching on the cloud is Netco~\cite{jalaparti2018netco}, a prefetching extension
integrated into the Hadoop Distributed File System (HDFS). Future data is prefetched based on two measures: 1) size of the
data to be processed and 2) task deadline.  Netco demonstrated superior performance compared to other file systems, which importantly motivates our study. However, it remains tightly bound to HDFS while cloud applications generally use different file systems. A more versatile solution is needed that would broadly apply to cloud data analysis and storage.

The present study focuses on neuroscience data that describe long-range
connections between different parts of the human brain, a central research topic in contemporary
neuroscience~\cite{bassett_network_2017}. The three-dimensional trajectory
of the major neural pathways, composed of millions of neuronal axons, are
inferred from measurements of diffusion MRI and processed using computational tractography algorithms. These
algorithms generate ``streamlines": 3D curves that approximate the
trajectories of the major pathways. A single human brain measurement may
contain several millions of these streamlines, with their coordinates
assessed at sub-millimeter resolution. Subsequent analyses of these
streamlines usually access
streamlines sequentially and entirely within the files in which they are stored. Such an access pattern creates an excellent opportunity for prefetching.

This paper investigates the benefits of prefetching for cloud-based processing of neuroscience streamlines. Through
both theoretical analysis and experimentation, we characterize the
speed-up provided by prefetching compared to sequential data transfers for
neuroscience data processing deployed on the Amazon Web Services cloud. More specifically, this paper makes the following contributions: 
\begin{itemize}
    \item Formalization and performance analysis of a ``rolling prefetch" data  scheme for cloud-based applications;
    \item Implementation based on the S3Fs Python library to access data on Amazon S3;
    \item Experimental evaluation in the Amazon cloud with a 500-GB dataset of streamlines derived from dMRI data.
\end{itemize}
% This is due to technical and scientific breakthroughs, enabling improved
% measurements that are providing more detailed views of the brain. But also
% thanks to sociotechnical trends that are  promoting increase sharing of
% neuroscientific data. For example, human brain imaging datasets are rapidly
% increasing in size. While many neuroimaging studies ranged from 10 -
% 100~GBs in size, newer datasets such as the
% BigBrain~\cite{amunts2013bigbrain}, UK Biobank~\cite{sudlow2015uk} and the
% Human Connectome Project (HCP)~\cite{van2013wu} are expected to reach up to
% Petabytes of data. Cloud platforms are commonly used to store this data in
% publicly-accessible object stores, as is the case for instance for \hcp
% data.

% \subsection{Cloud Infrastructure}
% The cloud gives users access various types of self-maintained services and hardware.
% This is a particularly attractive model to scientific groups who would not otherwise
% have the expertise nor the funds to maintain such infrastructure. Furthermore, the amount 
% of infrastructure is vast, supporting many users without significant queuing time in attaining the resources.
% Unlike High Performance Computing (HPC) clusters which may also provide access to various self-maintained
% hardware, users using the cloud as an Infrastructure as a Service (IaaS), can install whatever software is required
% on there.

\section{Materials and Methods}

Our implementation of prefetching, known as Rolling Prefetch, is a Python library implemented as a layer on top of \sfs. The code is available under the MIT license at \url{https://github.com/ValHayot/rollingprefetch}.
\sfs is a Python library, based on FSSpec, for interacting with directories and files located on Amazon S3.
To offset the cost of latency on S3, \sfs leverages the on-demand caching mechanisms provided by FSSpec.

Unlike \sfs which has distinct data transfer and compute phases, prefetched data transfers happen concurrently with compute (Figure~\ref{fig:prefetch}). In both cases, data is transferred from cloud storage in blocks of configurable size. While local I/O overheads occur during prefetch, the resulting overhead is in general minimal since memory disks are expected to be used for this purpose. Most importantly, prefetching requires knowledge of the application's data access pattern, such that transfers can be correctly anticipated. In our case, we assume that data is read sequentially, as is commonly the case with tractography data.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/prefetch_diagram.pdf}
\end{center}
\caption{Prefetching vs sequential transfers (\sfs)}
%https://docs.google.com/drawings/d/1MV5dIdLyBrM_HJX-0jOsIKf90UD8wXE_G5Cnmd-mY-A/edit
\label{fig:prefetch}
\end{figure}

\subsection{Algorithm}

% These caching mechanisms include MMapCache, ReadAheadCache,
% BlockCache, BytesCache and AllBytes. AllBytes attempts to cache the entire file contents into memory upon data access.
% MMapCache creates a memory-mapped temporary file which is populated in units of blocks, based on
% the configurable \texttt{block\_size} parameter, as data is requested. For small sequential reads, the ReadAheadCache 
% can be used to load data in blocks at a time. The BytesCache is similar to the ReadAheadCache except it improves the
% efficiency of semi-random reads by also maintaining parts of the previous cached block in memory. BlockCache differs from
% the others in that it implements a least recently used (LRU) cache of max number of blocks. When memory is filled, the
% least recently used block is evicted and the next one is read. By default, \sfs uses the BytesCache with a block size of
% 5~MB.

Rolling Prefetch combines prefetching with data eviction. By evicting data post application processing, it ensures a reduced footprint on local storage. Rolling Prefetch consists of three threads: (1) the reading thread loads data blocks from local storage and marks them for eviction, (2) the prefetching thread transfers data blocks from cloud storage to local storage, and (3) the eviction threads deletes the blocks that have been marked for eviction.
% \valerie{cut read alg below}
% \begin{algorithm}
% \SetAlgoLined
%     \SetKwInOut{Input}{Input}
%     \SetKwInOut{Output}{Output}
%     \Input{$start$ the start position of the read;\\
%     $end$ the end position of the read;\\}
%     \Output{$output$ the file contents located between $start$ and $end$}

%     $not\_local \gets True$\;

%     $read\_length \gets 0$
%     $total\_read \gets end - start$

%     \While{$total\_read > read\_length$}{
%         \While{$not\_local$}{
%             $block, not\_local \gets check\_local(start)$\;
%         }

%         $curr\_end \gets max(block.end, end)$\;
%         $output \gets block.read(start, curr\_end)$\;

%         \If{$block.end < end$}{
%             $start \gets curr\_end$\;
%             $flag\_for\_deletion(block)$\;
%         }

%     }

% \caption{Read}\label{alg:read}
% \end{algorithm}

\subsubsection{Reading}

The reading thread is triggered by the application to read data directly from cache or to wait until it is prefetched, if not found in the cache. %(Algorithm~\ref{alg:read}).
By waiting for the data to be cached, we ensure that performance is comparable to \sfs in a worst case scenario. Furthermore, as all data reads
are considered to be sequential, if the current block is not in memory then it is the current block being prefetched. Whenever a prefetched block has been read fully, it is up to the
read function to flag it for deletion.
% This process involves renaming the file to contain
% a ``.todelete'' extension. Using this extension, the eviction thread can determine which
% files need to be deleted.

\subsubsection{Prefetching}

Throughout the lifetime of the application, the prefetching thread continuously requests blocks from cloud storage, so long as there remain blocks that have not been prefetched (Algorithm~\ref{alg:prefetch}). Each block will be written to an appropriate cache location while not exceeding user-defined space limits.

Initially, the \texttt{used} variable for all cache locations is set to zero and the \texttt{file\_index} variable
is set to point to the first file index in the sequential ordering. The algorithm then enters a while loop whose
termination is dictated by the main program.
% (i.e., if the file object is still open, fetch, otherwise, terminate thread).
If all files have been prefetched, the prefetching threads terminates regardless of the status of the main thread.
Within the loop, the algorithm iterates between all available cache locations in priority order. The amount of available storage space
remaining is computed, and if it is insufficient to write the next block, the algorithm iterates through the file list $total\_files$
($verify\_used()$) and queries the filesystem to determine which blocks have been evicted. Based on how many blocks have been
removed, we can update the amount of used space, and by extension, the amount of available space. Should sufficient space
be released, the next block is fetched from cloud storage and is written to local storage. If there is insufficient space, the
next local storage location is tried.

%%mention in discussion the risk of an infinite loop if insufficient cache location storage space is provided. maybe 
% need for timeout? -- will not mention!

\begin{algorithm}
\SetAlgoLined
    \SetKwInOut{Input}{Input}
    \Input{
    $fetch$ a shared variable indicating whether the main thread has terminated or not\\
    $total\_files$ the list of all the files to prefetch;\\
    $cache\_location$ list of paths to cache locations;\\
    $total$ total space available on prefetch cache;\\
    $total\_size$ cumulative sum of all prefetched files;\\
    $blocksize$ size of a block\\}

$used \gets 0$\;
$file\_index \gets 0$\;
\While{$fetch$}{
\ForEach{$cache\_location$}{
$available \leftarrow total - used$\;
\If{ $available < blocksize$}{
    $available \gets verify\_used()$\;
}
\If{ $available \geq blocksize$}{
    $fetch\_next\_block()$\;
    $used \leftarrow used + blocksize$\;
}
}
\If{$total\_size \leq offset$ or $file\_index + 1 < total\_files$}{
    $file\_index \gets file\_index + 1$\;
    $total\_size \gets sizeof(total\_files[file\_index])$\;
}
\ElseIf{$total\_size \leq offset$}{
    break;
}
}
\caption{Prefetching}\label{alg:prefetch}
\end{algorithm}

\subsubsection{Eviction}

Similar to prefetching, the eviction thread only terminates when the main thread has completed processing. %(Algorithm~\ref{alg:eviction}).
% However, there is no mechanism for it to break prior to the main thread indicating termination. This 
% is because we assume that the file object will be closed when all data has been read.
To avoid additional costs caused by continuously querying the filesystem to determine which files can be evicted,
we determine the names of all the blocks that should be evicted ($get\_all\_blocks$), and verify whether they exist
in the filesystem at time of removal. We then update the list to ensure that we do not attempt to remove this file more 
than once. Between each loop, we sleep for 5 seconds to ensure that sufficient time has elapsed between evictions.

The eviction thread ensures deletion of all remaining files prior to terminating.

% \valerie{cut eviction alg below}
% \begin{algorithm}
% \SetAlgoLined
%     \SetKwInOut{Input}{Input}
%     \Input{$fetch$ a shared variable indicating whether the main thread has terminated or not;\\
%     $total\_files$ a list files to prefetch;\\
%     $cache\_location$ a list of paths to cache locations;\\
%     $file\_sizes$ a list containing the size of each file;\\
%     $blocksize$ size of a block\\}

%     $all\_blocks \gets get\_all\_blocks()$\;

%     \While{fetch}{
%         \ForEach{$block$ in $all\_blocks$}{
%             \ForEach{$dir$ in $cache\_location$}{
%                 $removed \gets remove(block)$\;

%                 \If{$removed$}{
%                     $update\_list(all\_blocks, block)$\;
%                     break\;
%                 }

%             }
%         }
%         $sleep 5$\;
%     }
%     $remove\_remaining(all\_blocks)$

% \caption{Eviction}\label{alg:eviction}
% \end{algorithm}

\subsection{Performance analysis}

We consider the cost of processing a file using sequential transfers (\sfs) to be the sum of three components (Equation~\ref{eq:s3fs}):
1) latency-associated costs, 2) bandwidth-associated costs and 3) application compute time.
The cost of latency is experienced every time a chunk of data is requested from cloud storage. That is, if we read a full file in a single call to cloud storage, we will only pay for latency once. If we issue
multiple calls to obtain file data on cloud storage, we will pay latency for each call. In contrast,
bandwidth is not affected by the number of calls. Whether we request all the data at once
or in chunks, bandwidth remains fixed, and time to transfer the data is
the quotient between the total amount of data transferred and the bandwidth. Compute time is assumed to be proportional to data size, as is frequently the case in neuroimaging.

    \begin{equation}
T_{\mathrm{seq}} = n_{b}l_c + \frac{f}{b_{cr}} +  cf,
\label{eq:s3fs}
\end{equation}
where $n_{b}$ is the number of data blocks, $f$ is the total size of the file to transfer, $l_c$ is the cloud latency, $b_{cr}$ is the cloud read bandwidth, and $c$ is the compute time per byte consumed.

Rolling Prefetch contrasts Equation~\ref{eq:s3fs} in that the compute and data transfer times mask one another (Equation~\ref{eq:prefetch}). However, Rolling Prefetch has a slightly higher performance penalty than that of sequential transfers when there is no compute, due to reading and writing the data to local storage. Furthermore, we
must consider the initial read from cloud storage, where no compute can occur concurrently, and the last compute, where no data transfer can occur concurrently.

\begin{equation}
T_{\mathrm{pf}} = T_{\mathrm{cloud}} + (n_b-1)\max\left(T_{\mathrm{cloud}}, T_{\mathrm{comp}}\right) + T_{\mathrm{comp}} \label{eq:prefetch}
\end{equation}
where $T_{\mathrm{cloud}}$ is the time to download a block from the cloud and write it to local storage, and $T_{\mathrm{comp}}$ is the time to read a block from local storage and process it:
\begin{eqnarray*}
T_{\mathrm{cloud}} &=& \overbrace{l_{c} + \frac{f}{b_{cr}n_{b}}}^{\mathrm{cloud\ read}} + \overbrace{l_{l} + \frac{f}{b_{lw}n_{b}}}^{\mathrm{local\ write}}, \\
T_{\mathrm{comp}} &=& \underbrace{l_l+\frac{f}{b_{lr}n_b}}_{\mathrm{local\ read}} + \underbrace{\frac{cf}{n_b}}_{\mathrm{compute}},
\end{eqnarray*}
where $l_l$ is the latency of local storage, $b_{lw}$ is the write bandwidth to local storage, and $b_{lr}$ is the read bandwidth to local storage.

This simple model provides several insights. First, if we neglect local transfers ($l_l$=0, $b_{lw}$=$b_{lr}$=+$\infty$), a reasonable assumption when local storage is in-memory, then we have:
\begin{equation*}
T_{\mathrm{seq}} = T_{\mathrm{pf}} + (n_b-1)\min\left(T_{\mathrm{cloud}},T_{\mathrm{comp}}\right),
\end{equation*}
and therefore the speed-up provided by Rolling Prefetch compared to sequential transfers is:
\begin{equation}
S=\frac{T_{\mathrm{seq}}}{T_{\mathrm{pf}}}=1+(n_b-1)\frac{\min\left(T_{\mathrm{cloud}},T_{\mathrm{comp}}\right)}{T_{\mathrm{pf}}} < 2 
\label{eq:speed-up}
\end{equation}
Rolling Prefetch is therefore expected to provide a speed-up of at most 2x. This upper bound is approached when $T_{\mathrm{cloud}} \approx T_{\mathrm{comp}}$, which requires that cloud transfer and compute times are of similar magnitude.

In sequential transfers, using a single block ($n_b=1$) leads to the shortest transfer time. In practice, 
block size is of course constrained by available memory. In contrast, with Rolling Prefetch, an optimal block size $\hat n_b$ exists under the reasonable assumption that $l_l \ll l_c$:
\begin{equation}
    \hat n_b = \sqrt{\frac{cf}{l_c}}
\end{equation}
It suggests that on a given cloud infrastructure, the number of blocks should increase when the application compute time or the total data size increases.

Finally, as the number of blocks increases, $T_{\mathrm{seq}}$ and $T_{\mathrm{pf}}$ become equivalent to $n_bl_c$ and $n_b(l_c+l_l)$, respectively, resulting in parallel asymptote lines.

%\tristan{Processing time should be explained when you explain the problem in intro. The real question in this paper is whether opening (and processing) tractography files involves enough data transfers AND compute time for prefetching to be useful. If one is negligible compared to the other, speed-up will be negligible too.}
% \ariel{Added a suggested subsection describing the dataset}
% \valerie{cut entire implementation subsection} \tristan{I moved the description of S3FS here, you could summarize the paragraphs on your implementation into 1}
% \subsection{Implementation}

% Rolling prefetch is a Python library implemented as a layer on top of \sfs. 
% \sfs is a Python library, based on FSSpec, for interacting with directories and files located on Amazon S3.
% To offset the cost of latency on S3, \sfs leverages the on-demand caching mechanisms provided by FSSpec.
% These caching mechanisms include MMapCache, ReadAheadCache,
% BlockCache, BytesCache and AllBytes. AllBytes attempts to cache the entire file contents into memory upon data access.
% MMapCache creates a memory-mapped temporary file which is populated in units of blocks, based on
% the configurable \texttt{block\_size} parameter, as data is requested. For small sequential reads, the ReadAheadCache 
% can be used to load data in blocks at a time. The BytesCache is similar to the ReadAheadCache except it improves the
% efficiency of semi-random reads by also maintaining parts of the previous cached block in memory. BlockCache differs from
% the others in that it implements a least recently used (LRU) cache of max number of blocks. When memory is filled, the
% least recently used block is evicted and the next one is read. By default, \sfs uses the BytesCache with a block size of
% 5~MB.

% \subsection{Implementation details}

% Inheriting functionality from \sfs permitted us to focus on prefetching rather
% that the logic of communicating directly with the S3 interface. However, we have modified some of the S3Fs functions, settings and functionality.

% \begin{sloppypar}
% The implementation can be split up into two classes: \texttt{S3PrefetchFileSystem} and 
% \texttt{S3PrefetchFile}. \texttt{S3PrefetchFileSystem} inherits from \sfs' \texttt{S3FileSystem}, and is nearly identical to the S3FileSystem
% class, with the exception that it overrides S3FileSystem's call to \texttt{open} to create
% and return an \texttt{S3PrefetchFile} object.
% \end{sloppypar}

% % Aside from reimplementing the \texttt{read} function, S3File is responsible for starting and stopping the prefetch and evictions threads. Both the prefetch and eviction threads are instatiated in the constructor of an S3PrefetchFile.
% % The threads are then terminated via the status
% % of a shared object (i.e. $self.fetch$) between the threads. 
% % The most important difference between \texttt{S3File} and \texttt{S3PrefetchFile} is that we have disabled \sfs caching in our library, since it is embedded in prefetching.

% Since it is common for large files to be broken down into datasets of smaller files, we wanted the ability to leverage prefetching in this case.
% As a result, a call to open accepts a list of file paths rather than just a single path. In the case where there needs to be a header,
% the first file in the provided list must point to the header region. If all the other files contain a header that needs to be skipped, there is
% a parameter, known as \texttt{header\_bytes} that can be passed to \texttt{open} to specify the size of the header to skip in subsequent files.
% Currently, it is not possible to prefetch unrelated files.

% % \begin{sloppypar}
% % The library fixes the open mode to be ``read bytes'', as \texttt{S3PrefetchFile} is only intended to read files, and currently only functions with binary files. To be able to write or append to files, the \texttt{open} of the parent class of \texttt{S3PrefetchFileSystem}
% % will need to be used.
% % \end{sloppypar}

% Prefetching should be enabled to write to any available local filesystem. As a result, we have added the \texttt{prefetch\_storage} parameter
% to the initialization of an \texttt{S3PrefetchFile}. This parameter allows users to specify the desired caching locations and the amount
% of available space to use at each location. The cache directories are listed in terms of descending priority. That is, the first location should
% point a location on the fastest available local storage, e.g., tmpfs.

\subsection{Data}

To demonstrate the utility of Rolling Prefetch in neuroimaging, we analyzed data derived from a dMRI experiment conducted in a single subject. These data are special, in that a  Super-Resolution Hybrid Diffusion Imaging (HYDI) method was used to achieve an effective spatial resolution of 0.625 mm$^3$, much higher than the typical $\sim$2 mm$^3$ commonly used. The details of the HYDI method and parameters of data acquisition were previously described \cite{Elsaid2019-ez}. The source MRI data were 14.94 GB. The data were processed using a probabilistic tractography algorithm \cite{Berman2008-xg}, implemented as part of DIPY \cite{Garyfallidis2014-el} and accelerated using CUDA to work on GPU \cite{rokem2021gpu}. This  generated $\sim$498 GB of streamline data. Tractography streamlines are usually stored in the neuroscience-specific \texttt{.trk} file format. In this case, streamlines were sharded into 464 \texttt{.trk} files stored in an S3 bucket using the high-availability Standard storage class.

% \tristan{I moved the nibabel text here, it could be summarized in 1 paragraph}
% \ariel{ I revised the nibabel section into one paragraph. Below in comments is the original as I found it:}
 Each \texttt{.trk} file is comprised of a 1,000-byte header and a body of variable length consisting 
of a series of streamlines. Each streamline section contains \SI{4}{\byte} that denote the number of points in the streamline,
followed by a series of floating point values detailing each coordinate and ends with a series of values representing properties of the streamline. We used Nibabel, a Python library that reads and writes  neuroscience-specific data formats, to read these files. For \texttt{.trk} files, Nibabel can return individual
data points via a generator. This is known as ``lazy loading'' and can be turned on or off prior to reading the file. As a result of the data representation format in \texttt{.trk} files, Nibabel reads may incur significant overhead: because it issues a total of three read calls for each streamline in the file.
% , some of which can be very small, depending on the streamline.
In addition, an affine transform is stored together with the data, and used to bring coordinates from different measurements into register. For example, when Nibabel reads \texttt{.trk} files it automatically applies an affine transformation to the coordinates of each streamline stored in the file. This means that some amount of compute is always executed when data is read from file.

% As in many other research fields, neuroscience data is usually stored in field-specific file formats. For example, MR images are often stored in the Neuroimaging Informatics Technology Initiative (NIfTI) file format, with the \texttt{.nii}  extension. Streamlines from computational tractography are often stored in a file format based on the Trackvis software, with the \texttt{.trk} extension. Nibabel is a Python library that implements reading and writing of many neuroscience-specific data formats. It is highly used in Python-based analysis projects of neuroimaging data. For files that
% exceed available memory, it provides the ability to MemoryMap them, it the case of randomly accessed files, or it returns individual
% data points via a generator, for sequentially accessed files. This is known as ``lazy loading'' and can be turned on or off prior to reading
% the file. In the case of \texttt{.trk} files representing streamlines, as they are sequentially read, Nibabel uses a generator to return streamlines.

% As a result of the data representation format in \texttt{.trk} files, Nibabel may incur significant overheads when
% reading them. Each \texttt{.trk} file is comprised of a 1000 byte header and a body of variable length consisting 
% of a series of streamlines. Each streamline contains \SI{4}{\byte} that denote the number of points in the track,
% followed by a series of floating point values detailing each track point and ends with a series of values representing the properties of the track. This means that for each existing streamline, Nibabel issues a total
% of three read calls, some of which can be very small, depending on the streamline.

% One of the challenges of analyzing volumetric data from neuroimaging is that the origin, voxel size and orientation of the acquisition volume can vary between different acquisitions (e.g., different MRI contrasts). To correspond the data from these different acquisitions, affine transforms are stored together with the data, and the transforms can be applied to the coordinates of different data to bring them into register. For example, when Nibabel reads \texttt{.trk} files it automatically applies an affine transformation to the coordinates of each streamline stored in the file. This means that some amount of compute is always executed when data is read from file.

\subsection{Experiments}
To evaluate the performance Rolling Prefetch, we conducted 4 experiments: 1) varying
the number of files; 2) varying the blocksize; 3) parallel processing of files;  and 4) neuroimaging use-cases. For experiments 1-3, the pipeline consisted of reading the file with Nibabel. As Nibabel
performs a minimal amount of computing, Rolling Prefetch is believed to have an effect even here. Additional compute required
by real analysis should only increase the benefit of using prefetching.
% For all experiments we are using tractography
% files belonging to the HYDI dataset stored on S3.

\subsubsection{Varying the number of files}\label{exp:files}

We vary the number of \texttt{.trk} files read to determine when Rolling Prefetch will be favourable relative to \sfs. We expected that for small amount of data, there should be no significant discernible difference between \sfs and Rolling Prefetch. This is because compute time is
expected to be minimal and the blocksize large compared to the total dataset size. In other words, there
will be less opportunities to prefetch data with very small datasets, unless the block size is proportionally smaller,
in which case latency will be penalizing both reads from \sfs and Rolling Prefetch, extending processing time. With larger files, there will be more computation occurring during the reads, and therefore, more opportunities
to mask the data transfers within the compute.

We benchmarked the time it took to lazily read 1, 5, 10, 15, 20 and 25 files and extract the streamlines in Nibabel.
The total data size for each of the file increments are 1, 6, 12, 18 and \SI{24}{\gibi\byte},
respectively. The blocksize was set to \SI{64}{\mebi\byte} and the prefetch storage was located on \texttt{tmpfs} with
a limit of \SI{2}{\gibi\byte} of storage space. We performed 10 repetitions.

\subsubsection{Varying the blocksize}\label{exp:blocksize}
In this experiment, we aimed to quantify the effect of the number of blocks ($n_b$) on Rolling Prefetch and \sfs. According to our previous analysis, we expect that \sfs will reach its best performance with a single block, and that the runtime will linearly increase with $n_b$ from that point on. For Rolling Prefetch we expect the runtime to decrease to a minimum, and then increase linearly with $n_b$.

Block sizes of 8, 16, 32, 64, 128, 256, 512, 1024,
and \SI{2048}{\mebi\byte} were used to load the files into Nibabel and extract the streamlines.
% We suspect that \SI{8}{\mebi\byte} may be too small to have many computes spanning more than
% a single block, however, it will occur at a higher frequency then with larger blocks.
We use the 
largest-sized block (\SI{2048}{\mebi\byte}) to determine the
overhead of using the Rolling Prefetch algorithm, as no file in the HYDI dataset is larger than \SI{1.7}{\gibi\byte},
and therefore, there are no opportunities to prefetch.

Five files pertaining to the HYDI dataset were used to generate the results.
These files ranged from \SI{793}{\mebi\byte} to
\SI{1.5}{\gibi\byte} in size each. Since only Rolling Prefetch is capable of treating a list of files as a single file, a new
Nibabel object needed to be created for each file in the case of \sfs.

The experiment was executed on both \sfs and Rolling Prefetch for each block and repeated 10 times. Tmpfs was excluded from this
experiment as the \sfs blocksize would have no effect on its performance. Prefetched storage was set to be tmpfs and
configured to have \SI{2}{\gibi\byte} of available space such that the largest block size could fit in prefetch
storage.

\subsubsection{Parallel processing}\label{exp:parallel}

Since perceived
throughput may be affected by the number of threads used, we aimed to determine if the performance difference
between \sfs and Rolling Prefetch remains proportional to the amount of data being processed per thread.

If throughput is reduced by the number of active threads, it is expected that 
the cost of data transfers will outweigh the cost of computing, such that the difference between \sfs and Rolling Prefetch will
be very small. Furthermore, since S3 is a scalable distributed file system and local storage may not be (or may be more limited in terms of scalability), local storage
throughput is expected to be reduced significantly with the addition of multiple threads. 
% Should this be the case, the
% overheads specific to prefetching will become very costly.

We used a total of four parallel processes to evaluate the effects of parallel processing. We increased the number of
files per thread between each condition from 1-20, with the last condition loading
a total of \SI{108}{\gibi\byte} into Nibabel. The blocksize was set to \SI{64}{\mebi\byte} and we set the prefetch
storage location to be on \texttt{tmpfs}, with a limit of \SI{1}{\gibi\byte} of storage per process. 5 repetitions were
performed.

\subsubsection{Neuroimaging use cases}\label{exp:4}
Our previous experiments evaluate the value of Rolling Prefetch when the application performs the most
basic and necessary action: converting the data from binary to array representation.
% Although it is unusual
% for applications to load the data without performing any additional processing of any kind, this will give us an
% idea of the lower bound of speedups .
We expect that the benefits of Rolling Prefetch can be maximized with an increased compute time, due to greater opportunities to prefetch data during compute. However, we also expect that there is a 
peak ratio between data transfer time and compute time (data-to-compute ratio), where the benefits of Rolling Prefetch will be highest.
If compute time is significantly greater than data transfer, regardless of all the time saved by prefetching data, the
total runtime will predominantly be compute. This motivates further profiling to determine how to shard
large datasets, like the HYDI dataset, such that the benefits of Rolling Prefetch can be maximized.

Since the benefits of Rolling Prefetch may vary according to the ratio between data transfer and compute time, we have selected two use cases with varying computation times: 1) histogram distribution of streamline lengths; and 2) bundle recognition.

The distribution of streamline lengths is used in order to assess the performance of tractography algorithms. It may also aid in determining if the files are good candidates for compression. The histogram computation consists of loading each streamline within the dataset lazily, gathering the lengths of
each individual streamline and generating a histogram of 20 bins from their lengths.
% Unlike the segmentation
% application, this application is data-intensive.

The human brain is composed of multiple different functional regions, and these regions are connected through major anatomical pathways. One of the first steps in the analysis of streamline data from human brain dMRI data is to classify the streamlines according to their three-dimensional trajectory into these different pathways. This is referred to as ``bundle recognition"~\cite{garyfallidis2018recognition}. To demonstrate the benefits of Rolling Prefetch in this use-case, we ran a software pipeline that determines whether streamlines within the tractography file belong to either of two different bundles (the corticospinal tract, CST and the arcuate fasciculus, ARC), or to neither \cite{Kruper2021-bo}. Currently, the pipeline loads the data all at once (i.e. no lazy loading) and then performs the bundle recognition task. Thus, it is not possible to read only fragments of a large file which does not fit in memory, in this case, and as a consequence of the separation of data loading and processing, there will be no opportunities to keep the prefetched data transfers masked within compute operations.

We chose not to modify the pipeline, in order to determine what is the speedup provided
by Rolling Prefetch without additional optimization, and in order to be able to speculate on the possible benefits of adapting existing pipelines
to support larger-than-memory files and allowing reads and computations to occur together.

Varying parameters were used for the bundle recognition and
histogram application. For instance, recognition was
executed on two different compute instances (c5.9xlarge and r5.4xlarge). The experiment executed on the c5.9xlarge instance consisted of a \SI{1}{\gibi\byte} file with a \SI{64}{\mebi\byte} blocksize. The histogram experiment executed on the r5.4xlarge instance used 10 files totalling \SI{12}{\gibi\byte} and used a \SI{32}{\mebi\byte} blocksize.

We also considered the speedups obtained from
bundle recognition in smaller files. We split up the 
previous \SI{1}{\gibi\byte} file and split it up into
9 shards containing the same number of streamlines. File sizes ranged from \SI{73}{\gibi\byte} to \SI{165}{\mebi\byte}. We compared the speedup provided by Rolling Prefetch on a \SI{165}{\mebi\byte} shard to
the processing of all 9 shards. These experiments were
run exclusively on the r5.4xlarge instance with 10 repetitions.

To compare the benefits of Rolling Prefetch with different data-to-compute ratios, we executed the histogram computation and the bundle recognition algorithm on
an r5.4xlarge instance using the \SI{1}{\gibi\byte} file. Blocksize in both cases was fixed at \SI{32}{\mebi\byte}. 5 repetitions were performed.

In all cases, \SI{2}{\gibi\byte} of cache storage was allocated to Rolling Prefetch.

\subsection{Infrastructure}

For the first three experiments, we used an Amazon EC2 t2.xlarge instance with 4 vCPUs and \SI{16}{\gibi\byte} of RAM with a
maximum of \SI{7.8}{\gibi\byte} of RAM of tmpfs space. The compute instance was configured to use Red Hat Enterprise
Linux 8.3 with kernel version 4.18.0. This instance was hosted in availability zone \texttt{us-west-2a}.
The experiments all used Python version 3.9.0 with version 0.5.2 of \sfs and Nibabel version 3.2.1.

The scientific applications required significantly more memory than available on t2.xlarge instances. As a result, the scientific applications were executed on a c5.9xlarge instance and an r5.4xlarge configured identically
to the t2.xlarge instance, but located in the \texttt{us-west-2d} availability zone. c5.9xlarge instances consist of 64 vCPUs and have \SI{72}{\gibi\byte} of available memory, whereas the r5.4xlarge instance consists of 16 vCPUs and \SI{128}{\gibi\byte} of available memory.

The data used for all experiments originated from the HYDI tractography dataset stored on the Amazon S3 object store. 
This dataset is broken down into 464 files ranging from \SI{700}{\mebi\byte} to \SI{1.7}{\gibi\byte} each. This dataset
was also located in the \texttt{us-west-2} region.

To get an idea of the cost of data transfers, we measured the time it would take to read data sequentially from S3 to the t2.xLarge instance. We varied the filesize from \SI{1}{\kibi\byte} to \SI{4}{\gibi\byte} and measure the time it took to read the files from
a \texttt{us-west-2} S3 bucket and from tmpfs. This method was favoured over more standard benchmarking libraries to ensure the
two filesystems were benchmarked in the same way. Results were then compared to benchmarking results produced by
the \href{https://github.com/dvassallo/s3-benchmark}{s3-benchmark} library to ensure validity of our benchmarking
method. The results can be seen in Table~\ref{table:benchmarks}.

%%tristan do i keep this if our model is just a theoretical eval? maybe it should be placed where the model
%% is instead?
\begin{table}
\caption{Measured latency and bandwidth obtained from reading files ranging from \SI{1}{\kibi\byte} to \SI{4}{\gibi\byte} on a t2.xLarge instance reading from an S3 bucket in the same region.}
\centering
\begin{tabular}{| c | c | c| }
\cline{2-3}
  \multicolumn{1}{c|}{}& S3 & memory \\ 
  \hline
 bandwidth (MB/s) & 91 & 2221 \\  
 latency (s) & 0.1 & $\num{1.6e-6}$ \\
 \hline
\end{tabular}

\label{table:benchmarks}
\end{table}

\section{Results}
\subsection{Varying the file size}
As we increase the file size we observe that the disparity between \sfs and
Rolling Prefetch increases (Figure~\ref{fig:filesize}), with Rolling Prefetch significantly outpacing ($\sim$1.7x faster) 
\sfs at \SI{26}{\gibi\byte} (25 files). This is expected as per our theoretical evaluation. With a large amount of blocks, the pipeline duration is determined by the maximum between the time it takes to
compute the data $T_{comp}$ and the time it takes to transfer the data $T_{cloud}$. On the other hand, \sfs runtime is the sum of the time it takes to transfer the data and the compute. In this case, the compute time is large enough such that using Rolling Prefetch significantly reduces runtime with large amounts of data.

As the size of the data decreases, the benefit of using Rolling Prefetch also decreases. 
% This too is expected, as there is proportionally less compute occurring.
In this particular use case, S3 data transfer time is much larger
than that of compute time, thus the only time we can save is that of $T_{comp}$. $T_{comp}$ like ${T_{cloud}}$ is
proportional to the number of blocks used, and thus the amount of time saved increases with file size. In the 
worst case, Rolling Prefetch performance equals that of \sfs.

\subsection{Parallelized workloads}
% \tristan{This section refers to file numbers while figure is in GiB}
When we parallelize the reads up to four concurrent processes (Figure~\ref{fig:parallel}), we notice that the trends are still consistent
despite increased contention on local filesystems. This is likely due to the fact that \texttt{tmpfs} speeds
would still be sufficiently high even with the increased contention caused by having at minimum 4 and 
potentially even up to 8 concurrent threads accessing the filesystem at the same time. We speculate that the same 
pipeline running with data read from a single disk would have had worse performance
due to its lack of scalability.

The maximum speedup achieved, on average, during this experiment was 1.8$\times$ at 5.8~GiB. The average speedup
was around 1.5~x altogether, with the minimum average speedup being 1.37~x at 18~GB. Due to the high
variability in results within both Rolling Prefetch and \sfs, we expect that these speedup differences may be a
result of bandwidth and latency variability across runs. We do not expect that conditions are more favourable
for reading the data with four threads processing 5.8~GiB. Since they are processing the
same amount of data concurrently, the overall speedup should be constant.

\subsection{Varying the block size}%% some parts may be better suited for discussion

At a very large number of blocks, the performance of both \sfs and Rolling Prefetch is degraded as a result of 
increased impacts of both S3 storage latency and, in the case of Rolling Prefetch, also local storage latency (Figure~\ref{fig:blocksize}).  \sfs starts to surpass the speed of Rolling Prefetch, although not substantially (approx. 1.03~X) at 747 blocks.  Furthermore, while we read from S3
at certain blocksizes, the block size on local storage is dictated by filesystem and kernel readahead 
size. In our specific case, the data would be read from tmpfs, and thus we do not see a significant added cost
to Rolling Prefetch. It could be inferred that the overhead of Rolling Prefetch may be more significant with a large
number of blocks due to Nibabel's access patterns, which consist of many small reads.

Conversely, when decreasing the number of blocks we start to see an increase in performance in both \sfs and Rolling Prefetch. This
is because both benefit from a decreased latency to access data on S3, by generally making fewer read calls to it. Rolling Prefetch is found to be faster than \sfs with a peak speedup of 1.24~X at 186 blocks (i.e. 
\SI{32}{\gibi\byte} blocks). The speedup obtained from Rolling Prefetch did not vary significantly between 23 and
186 blocks, averaging at a speedup of approximately 1.2~X. Prefetching starts to outpace \sfs a bit below
747 blocks as the cost of local latency has diminished sufficiently with the larger blocks. As the blocksize
increases, Rolling Prefetch is able to begin maskinf S3 latency and bandwidth within its compute.
% The speedup being
% proportional to how many more blocks can be read within the computation.
\sfs and Rolling Prefetch performance converges
once more as the number of blocks decrease and the block size increases, placing more weight disk bandwidth
than latency. \sfs exceeds the performance of Rolling Prefetch at a single block, where no actual prefetching can take
place and we pay the cost of implementation overheads, such as writing the data to local storage 
and reading it from local storage rather than directly from memory.

Generally speaking, variations in blocksize did not lead to major speedups between Rolling Prefetch and \sfs. Since the
total files size and compute time was fixed, there was a fixed upper bound to how many data transfers could
be masked by compute time. By minimizing the amount of latency through a reduction in the number of blocks, the 
overall application time was shortened, in both \sfs and Rolling Prefetch, and thus a larger percentage of the total 
runtime could be masked by compute.

\subsection{Neuroimaging use-cases}

The results obtained from our neuroimaging use cases indicate that, in all
cases, the use of Rolling Prefetch can speed up analyses (Figure~\ref{fig:speedup}).
The extent of this speedup is, however, dependent on a number of
factors, such as instance type, application and number of shards.

The greatest observed speedup with Rolling Prefetch was found when executing the bundle recognition application
with a greater number of shards (1.64$\times$ faster). This result echoes what was observed in Figure~\ref{fig:filesize} with even a more compute-intensive application like 
bundle recognition. We do not, however, observe a speedup with a single shard, as the size of the shard was so small it did not incur many reads from S3.

Due to the compute-intensive nature of the bundle recognition pipeline, it comes as no
surprise that overall speedup in the unsharded case is minimal (1.14$\times$). The ratio of data-to-compute was found to be approximately 1/7, where the compute took, on average around 9000s. Thus, even if Rolling Prefetch significantly speeds up reads, the data transfer time is
minimal to the overall compute time of the application. With a highly data-intensive
workflow, such as a histogram computation, we observe that the speedup is more
significant with a speedup of 1.5$\times$. Although our model dictates that the upper bound
of speedup that can be obtained with Rolling Prefetch is 2$\times$, the observed speedups
obtained by these two applications never reach that boundary. A possible explanation may be that the data-to-compute ratio of histogram computation was too large to mask many of
the reads within the compute, whereas the data-to-compute ratio was too small in the segmentation application, such that any speedup obtained
through the Rolling Prefetch algorithm was offset by the computation time. The ideal
algorithm for Rolling Prefetch would lie somewhere between the data and computational
requirements of the histogram and segmentation applications.

Although the execution parameters differed between the two instance types thus not
making them directly comparable, we observed a greater speedup on the c5.9xlarge
(1.6$\times$) than on the r5.4xlarge (1.1$\times$) instance. We suspect that the speedup obtained
on the c5.9xlarge instance was relatively high as a result of the increased parallelism
obtained from the increased number of CPUs, resulting in a more data-intensive application. While the r5.4xlarge instance did process a larger number of files, 
suggesting that speedup should be greater, the actual file sizes varied, and with
it, the streamline features, potentially resulting in a more compute-intensive
execution further exacerbated by the decrease in available vCPUs.

% old paragraph on segmentation
% Results on the segmentation pipeline demonstrate that prefetching provides a substantial speedup over \sfs. The
% average speedup obtained, as can be seen in Figure~\ref{fig:segmentation}, was approximately 1.5~x. This performance gain is quite substantial despite data loading
% and processing being distinct stages within the pipeline. It is expected that we can obtain an even
% greater performance boost with prefetching if data loading and processing occur together, obtaining a speedup
% closer to the limit of 2~X. Furthermore, due to the memory requirements of the pipeline, enabling large datasets
% to be loaded lazily with prefetching and Nibabel may reduce the memory demands of the application, enabling more
% cost-effective instances to be selected.

% Likewise, as per the results obtained in Experiment~\label{exp:filesize}, we expect the performance gain to
% increase with files size. We also expect this pipeline to adapt well to parallelism (as per Experiment~\label{exp:parallel}, up to the point until strain on the local storage due to filesystem contention 
% outweighs that of contention on S3.

\begin{figure}
\begin{center}
\includegraphics[height=160pt]{figures/filesize.pdf}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-10pt}
\caption{Runtime performance of reading subsets of the HYDI dataset stored on Amazon S3 into Nibabel using \sfs and Rolling Prefetch on an Amazon EC2 t2.xlarge instance. Blocksize was set to \SI{64}{\mebi\byte} on both \sfs and
Rolling Prefetch. Prefetch cache consisted of \SI{2}{\gibi\byte} tmpfs storage.}
\label{fig:filesize}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/parallel.pdf}
\setlength{\abovecaptionskip}{-10pt}
\setlength{\belowcaptionskip}{-10pt}
\caption{Runtime performance of reading subsets of the HYDI dataset stored on Amazon S3 into Nibabel in parallel using 4
\sfs and Rolling Prefetch processes. Blocksize was set to \SI{64}{\mebi\byte} on both \sfs and Rolling Prefetch.
Prefetch cache consisted of \SI{1}{\gibi\byte} tmpfs storage.}
\label{fig:parallel}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/blocksize.pdf}
\setlength{\abovecaptionskip}{-10pt}
 \setlength{\belowcaptionskip}{-20pt}
\caption{Runtime performance of reading a \SI{6}{\gibi\byte} subset of the HYDI dataset stored on Amazon S3 into Nibabel using
\sfs and Rolling Prefetch with various block size configurations on an Amazon EC2 t2.xlarge instance. Prefetch cache consisted of  \SI{2}{\gibi\byte} tmpfs storage.}
\label{fig:blocksize}
\end{center}
\end{figure}

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figures/segmentation.pdf}
% \caption{Runtime performance of the Segmentation pipeline operating on a \SI{1}{\gibi\byte} S3 HYDI tractography file using \sfs and prefetch on an Amazon C5.9xlarge instance. Blocksize was set to be \SI{64}{\mebi\byte} on 
% both \sfs and prefetch. Prefetch was allotted \SI{2}{\gibi\byte} of storage on tmpfs.} 
% \label{fig:segmentation}
% \end{center}
% \end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/speedup.pdf}
\setlength{\abovecaptionskip}{-10pt}
\setlength{\belowcaptionskip}{-8pt}
\caption{Rolling prefetch speedup of the neuroimaging use-cases (histogram and bundle recognition) in various conditions. Experiments varying instance type and number of files were only performed with the bundle recognition pipeline.} 
\label{fig:speedup}
\end{center}
\end{figure}

\section{Discussion}

Our theoretical analysis and experimental results demonstrate that there is a substantial processing time gain to be
obtained from using Rolling Prefetch, particularly in the case of mixed workloads, where there is a significant
cost associated with time spent on compute and data transfers. This works well with typical use cases in neuroimaging,
where tasks vary widely ranging from very short tasks to long ones necessitating hours to complete and where datasets
are large enough to incur transfers of similar magnitudes. Moreover, to save time with data transfers, researchers
may opt to transfer their data over the compute instance, and perform the computation exclusively with
data stored locally. While this does effectively give the optimal performance during processing, researchers are left to manage
the data themselves. Since local storage on compute can become quite costly, researchers must decide between
processing only a subset of the data, paying hefty fees related to storing large amounts of data or 
incorporating data management logic directly into their workflows.

There are also natural limitations to
Rolling Prefetch. For instance, in the case of parallel workloads, in certain instances S3 would be
preferred to the available local storage. This is a consequence of the fact that S3 is designed to be scalable
and is capable of processing many requests simultaneously, whereas unless configured to do so, local storage
will not scale adequately to increased contention.
% Additionally, when blocks become significantly large with
% an equally large number of blocks, it is possible that rolling prefetch will be slower than \sfs. 
% -- commented out because cannot remember what i meant

\subsection{Benefits of Rolling Prefetch}

Rolling Prefetch is an added layer on top of \sfs that replaces its built-in cache strategies to intercept
application read calls and ensure that data is preloaded. The implementation ensures that filesystem space requirements are not exceeded during prefetching through limits set by the user. 
With the built-in eviction mechanism, users are not required to do any form of data management should they be processing datasets larger than available storage. Furthermore, the library allows configuration of multiple storage device as cache for prefetching. Currently,
files are written to specific storage devices based on available space and storage priority in the list.

With just a simple implementation performed on a computation based on reading, we have observed up to a 1.8$\times$ speedup and a maximum overhead of 1.03$\times$. These observed speedups were
obtained when we set the cache location to a \texttt{tmpfs} device and may naturally decrease with a slower device such as an SSD. In our particular case, the speedups meant saving 20~min of processing time
on loading nearly \SI{100}{\gibi\byte} of data with a maximum runtime of approximately 71~min. Moreover, this
was achieved on a small instance with only \SI{1}{\gibi\byte} of prefetch storage allotted to each parallel
process, indicating that we can achieve good speedups even when resources are constrained.

\subsection{Parallelism-related overheads}

While our experiments demonstrate that Rolling Prefetch continues to provide a performance improvement to parallel
applications running on S3, we do expect  performance to decrease if we continue to scale horizontally on
a single instance, or use a slower device as cache. Our implementation consists of two threads actively reading and writing to local storage. Each time the number of Rolling
Prefetch instances increase, we double the amount of threads writing to local storage. Although it is standard
to have multiple DIMMs on a single machine, it may
not be necessarily true of other storage devices. That being said, attached cloud volumes may
also be sufficiently scalable such that processing time remains unaffected by an increase in processes.

To reduce the load of prefetching data to local storage, we can add a third component to Rolling Prefetch that periodically tracks cache bandwidth. Using such a parameter, the algorithm could take into consideration filesystem 
load in addition to user-specified priority.

\subsection{Task granularity}
\sfs was designed to be used within distributed processing frameworks. With such frameworks, we must take into consideration task granularity and how that would affect
scheduling. Rolling prefetch becomes beneficial with larger files. Assuming a workload where the task is simply to load the streamlines, we would need a few
GBs of data to start noticing a speedup. In cases where resources are ample enough to run all tasks in
parallel and the amount of data processed by a task is minimal, Rolling Prefetch and \sfs would probably behave
similarly, with \sfs potentially performing a bit faster.

The risk with passing large amounts data to the library within a task is that it is less robust to failures as any failed task would have to resume from the beginning. This could
take a significant amount of time, depending on how much data had been processed by the task prior to failure.
% Making the tasks smaller would reduce the costs associated with task restart, but then the benefits of
% Rolling Prefetch would be limited.
It is understandable that implementations such as Netco are adaptations to
persistent filesystem, since then prefetching could work without limiting scheduler logic. For our 
implementation to be efficient within parallel processing frameworks, we would need to decouple it from
the task itself and allow it to communicate with the task scheduler to determine which files to prefetch and
where.

\section{Conclusion}

% Prefetching is a widely used method for minimizing the costs of computational workloads that involve multiple data reads.
% To determine the potential added value of prefetching to large-scale neuroimaging workloads, we quantified its
% speedup through analysis of the runtime performance of standard neuroscience data loading, and by embedding this idea in existing computational pipelines. To accomplish this, we developed Rolling Prefetch, a Python library that can
% prefetch sequentially-read binary files located on S3 as an extension of the \sfs library. We compared the performance of Rolling Prefetch with \sfs.
% While there have been
% implementations for prefetching on the cloud, this technique has not been used for the processing of neuroscience
% workloads on the cloud. However, this idea is becoming more relevant for neuroscience data analysis workloads, as contemporary neuroscience datasets continue to increase in size and analysis pipelines are becoming more data-intensive.

% Our theoretical analysis of prefetching concluded that Rolling Prefetch can provide up to a maximum of 2x speedup, 
% and is most effective when the time needed to compute is comparable to the time required to transfer the data.
% Furthermore, it was determined that while \sfs performed best given the least number of blocks, Rolling Prefetch 
% would perform best with an intermediary number of blocks, where the overheads of prefetching would become more
% significant at 1 block, when prefetching would at minimum do the same amount of work as \sfs, and at many blocks,
% where the latency of writing locally would become apparent.
% Overall the analysis states that it is better to
% slightly overestimate the number of blocks as the penalty of reading more blocks is proportional to the latency
% of the cloud ($l_{c}$).

% Our experimental evaluation supports the conclusions drawn from the theoretical evaluation. A peak speedup of
% 1.8x was attained when processing a total of 20 tractography files across four parallel threads. Good speedups
% were also obtained when increasing the number of files processed per task, obtaining a 1.7x speedup at 25 files.
% As determined by our theoretical evaluation, having too few blocks and too many blocks penalized the performance
% of Rolling Prefetch due to overheads incurred by the prefetching process.

Overall, we conclude that Rolling Prefetch would be a valuable addition to large-scale neuroimaging pipeline executing
on the cloud, particularly in instances where data transfer and compute time are similar. 
% However, our implementation still has limitations that may not make it favorable to distributed
% processing. 
Future work should focus on the following improvements: The loading part of Rolling Prefetch should be performed outside of task themselves to ensure that it does not interrupt any form of task scheduling and to avoid tasks which take too long to restart if lost. 
The library should also make sure to be able to communicate with the schedulers to help determine where tasks
should be scheduled give the location of the data.

\section{Acknowledgments}

This research was supported by NIH grant 1RF1MH121868-01 and by cloud compute credits from Microsoft Azure. This research was also supported by the Canada Research Chairs program, and by the Natural Sciences and Engineering Research Council of Canada.

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{plain}
\bibliography{biblio}

\end{document}
\endinput
%%
%% End of file `sample-authordraft.tex'.

### 2021-08-23 paper/paper.tex
minor paper updates with figure generation script

@@
 Rolling Prefetch extends \sfs, a Python library exposing AWS S3 functionality via a file object, to add prefetch capabilities. 
-In measured analysis performance of a 500~GB brain connectivity dataset stored on S3, we found that prefetching provides significant speed-ups of up to 1.8$\times$, even in applications consisting entirely of data loading.
+In measured analysis performance of a 500~GB brain connectivity dataset stored on S3, we found that prefetching provides significant speed-ups of up to 1.86$\times$, even in applications consisting entirely of data loading.
 The observed speed-up values are consistent with our theoretical analysis.
@@
 a total of \SI{108}{\gibi\byte} into Nibabel. The blocksize was set to \SI{64}{\mebi\byte} and we set the prefetch
-storage location to be on \texttt{tmpfs}, with a limit of \SI{1}{\gibi\byte} of storage per process. 5 repetitions were
+storage location to be on \texttt{tmpfs}, with a limit of \SI{1}{\gibi\byte} of storage per process. 10 repetitions were
 performed.
@@
 Rolling Prefetch increases (Figure~\ref{fig:filesize}), with Rolling Prefetch significantly outpacing ($\sim$1.7x faster) 
-\sfs at \SI{26}{\gibi\byte} (25 files).
+\sfs at \SI{31.2}{\gibi\byte} (25 files).
 This is expected as per our theoretical evaluation.
@@
 
-\subsection{Parallelized workloads}
-% \tristan{This section refers to file numbers while figure is in GiB}
-When we parallelize the reads up to four concurrent processes (Figure~\ref{fig:parallel}), we notice that the trends are still consistent
-despite increased contention on local filesystems. This is likely due to the fact that \texttt{tmpfs} speeds
-would still be sufficiently high even with the increased contention caused by having at minimum 4 and 
-potentially even up to 8 concurrent threads accessing the filesystem at the same time. We speculate that the same 
-pipeline running with data read from a single disk would have had worse performance
-due to its lack of scalability.
-
-The maximum speedup achieved, on average, during this experiment was 1.8$\times$ at 5.8~GiB. The average speedup
-was around 1.5~x altogether, with the minimum average speedup being 1.37~x at 18~GB. Due to the high
-variability in results within both Rolling Prefetch and \sfs, we expect that these speedup differences may be a
-result of bandwidth and latency variability across runs. We do not expect that conditions are more favourable
-for reading the data with four threads processing 5.8~GiB. Since they are processing the
-same amount of data concurrently, the overall speedup should be constant.
-
 \subsection{Varying the block size}%% some parts may be better suited for discussion
 
-At a very large number of blocks, the performance of both \sfs and Rolling Prefetch is degraded as a result of 
-increased impacts of both S3 storage latency and, in the case of Rolling Prefetch, also local storage latency (Figure~\ref{fig:blocksize}).
+At a very large number of blocks, the performance of both \sfs and Rolling Prefetch is degraded due to
+increased impacts of both S3 storage latency and additionally, in the case of Rolling Prefetch, local storage latency (Figure~\ref{fig:blocksize}).
 \sfs starts to surpass the speed of Rolling Prefetch, although not substantially (approx.
-1.03~X) at 747 blocks.
+1.03~X) at 748 blocks.
 Furthermore, while we read from S3
@@
 Conversely, when decreasing the number of blocks we start to see an increase in performance in both \sfs and Rolling Prefetch. This
-is because both benefit from a decreased latency to access data on S3, by generally making fewer read calls to it.
-Rolling Prefetch is found to be faster than \sfs with a peak speedup of 1.24~X at 186 blocks (i.e.
-\SI{32}{\gibi\byte} blocks). The speedup obtained from Rolling Prefetch did not vary significantly between 23 and
-186 blocks, averaging at a speedup of approximately 1.2~X. Prefetching starts to outpace \sfs a bit below
-747 blocks as the cost of local latency has diminished sufficiently with the larger blocks. As the blocksize
-increases, Rolling Prefetch is able to begin maskinf S3 latency and bandwidth within its compute.
+is because both benefit from decreased latency to access data on S3, by generally making fewer read calls to it.
+Rolling Prefetch is found to be faster than \sfs with a peak speedup of 1.24~X at 187 blocks (i.e.
+\SI{32}{\mebi\byte} blocks). The speedup obtained from Rolling Prefetch did not vary significantly between 24 and
+187 blocks, averaging at a speedup of approximately 1.2~X. Rolling Prefetch outpaces \sfs at less than
+748 blocks as the cost of local latency has diminished sufficiently with the larger blocks. As the blocksize
+increases, Rolling Prefetch is able to begin masking S3 latency and bandwidth within its compute.
 % The speedup being
 % proportional to how many more blocks can be read within the computation.
-\sfs and Rolling Prefetch performance converges
-once more as the number of blocks decrease and the block size increases, placing more weight disk bandwidth
+\sfs and Rolling Prefetch performance converges again when the blocks reach a certain size, placing more weight disk bandwidth
 than latency. \sfs exceeds the performance of Rolling Prefetch at a single block, where no actual prefetching can take
@@
 runtime could be masked by compute.
+
+\subsection{Parallelized workloads}
+% \tristan{This section refers to file numbers while figure is in GiB}
+When we parallelize the reads up to four concurrent processes (Figure~\ref{fig:parallel}), we notice that the trends are still consistent
+despite increased contention on local filesystems. This is likely due to the fact that \texttt{tmpfs} speeds
+would remain sufficiently high even with the increased contention caused by having at minimum 4 and 
+potentially even up to 8 concurrent threads accessing the filesystem at the same time. We speculate that the same 
+pipeline running with data read from a single disk would have had worse performance
+due to its lack of scalability.
+
+The maximum speedup achieved, on average, during this experiment was 1.86$\times$ at 24.2~GiB. The average speedup
+was around 1.52~x altogether, with the minimum average speedup being 1.37$\times$ at 80.6~GB. Due to the high
+variability in results within both Rolling Prefetch and \sfs, we expect that these speedup differences may be a
+result of bandwidth and latency variability across runs. We do not expect that conditions are more favourable
+for reading the data with four threads processing 24.2~GiB; since they are processing the
+same amount of data concurrently, the overall speedup should be constant.
 

### 2021-08-24 paper/paper.tex
final mod

@@
 We benchmarked the time it took to lazily read 1, 5, 10, 15, 20 and 25 files and extract the streamlines in Nibabel.
-The total data size for each of the file increments are 1, 6, 12, 18 and \SI{24}{\gibi\byte},
+The total data size for each of the file increments are 1.1, 5.8, 11.9, 18.0, 24.2 and \SI{31.2}{\gibi\byte},
 respectively. The blocksize was set to \SI{64}{\mebi\byte} and the prefetch storage was located on \texttt{tmpfs} with
@@
 According to our previous analysis, we expect that \sfs will reach its best performance with a single block, and that the runtime will linearly increase with $n_b$ from that point on.
-For Rolling Prefetch we expect the runtime to decrease to a minimum, and then increase linearly with $n_b$.
+For Rolling Prefetch, we expect the runtime to decrease to a minimum, and then increase linearly with $n_b$.
 

### 2021-10-01 paper/paper.tex
addressing comments

@@
 
+Since Rolling Prefetch is implemented as a layer atop \sfs, it can easily replace an existing call to S3Fs by calling \texttt{S3PrefetchFileSystem} instead of \texttt{S3FileSystem}.
+
 \begin{figure}
@@
 
-Rolling Prefetch combines prefetching with data eviction.
-By evicting data post application processing, it ensures a reduced footprint on local storage.
+Rolling Prefetch combines prefetching with data eviction. By evicting data post application processing, it ensures a reduced footprint on the local storage leveraged as cache by Rolling Prefetch
+(e.g.
+tmpfs, SSD, HDD).
 Rolling Prefetch consists of three threads: (1) the reading thread loads data blocks from local storage and marks them for eviction, (2) the prefetching thread transfers data blocks from cloud storage to local storage, and (3) the eviction threads deletes the blocks that have been marked for eviction.
@@
 To evaluate the performance Rolling Prefetch, we conducted 4 experiments: 1) varying
-the number of files; 2) varying the blocksize; 3) parallel processing of files;  and 4) neuroimaging use-cases. For experiments 1-3, the pipeline consisted of reading the file with Nibabel. As Nibabel
+the number of files; 2) varying the blocksize; 3) parallel processing of files;  and 4) neuroimaging use-cases.
+For experiments 1-3, the pipeline consisted of reading the file with Nibabel. As Nibabel
 performs a minimal amount of computing, Rolling Prefetch is believed to have an effect even here. Additional compute required
@@
 
+Our results show that significant speedup can be gained even in instances where compute is rather short. While our results
+pertain to the applicability of Rolling Prefetch to neuroimaging, similar speedups can be expected from other kinds of sequentially read data processing,
+such as a filter on a very large file.
+
 \subsection{Parallelism-related overheads}

### 2021-10-04 paper/paper.tex
figure fix

--- 2021-08-24 paper/paper.tex
@@
 %https://docs.google.com/drawings/d/1MV5dIdLyBrM_HJX-0jOsIKf90UD8wXE_G5Cnmd-mY-A/edit
+\vspace{-1.5em}
 \label{fig:prefetch}
@@
 
-Rolling Prefetch combines prefetching with data eviction.
-By evicting data post application processing, it ensures a reduced footprint on local storage.
+Rolling Prefetch combines prefetching with data eviction. By evicting processed data, it ensures a reduced footprint on the local storage leveraged as cache
+(e.g.
+tmpfs, SSD, HDD).
 Rolling Prefetch consists of three threads: (1) the reading thread loads data blocks from local storage and marks them for eviction, (2) the prefetching thread transfers data blocks from cloud storage to local storage, and (3) the eviction threads deletes the blocks that have been marked for eviction.
@@
 \end{tabular}
-
+\vspace{-1.5em}
 \label{table:benchmarks}
@@
 \begin{center}
+    \vspace{-0.6em}
 \includegraphics[height=160pt]{figures/filesize.pdf}
 \setlength{\abovecaptionskip}{0pt}
-\setlength{\belowcaptionskip}{-10pt}
+\setlength{\belowcaptionskip}{-30pt}
 \caption{Runtime performance of reading subsets of the HYDI dataset stored on Amazon S3 into Nibabel using \sfs and Rolling Prefetch on an Amazon EC2 t2.xlarge instance.
@@
 Rolling Prefetch. Prefetch cache consisted of \SI{2}{\gibi\byte} tmpfs storage.}
+\vspace{-5.53em}
 \label{fig:filesize}
@@
 Prefetch cache consisted of \SI{1}{\gibi\byte} tmpfs storage.}
+\vspace{-5.5em}
 \label{fig:parallel}
@@
 \sfs and Rolling Prefetch with various block size configurations on an Amazon EC2 t2.xlarge instance. Prefetch cache consisted of  \SI{2}{\gibi\byte} tmpfs storage.}
+\vspace{-1.5em}
 \label{fig:blocksize}
@@
 Experiments varying instance type and number of files were only performed with the bundle recognition pipeline.}
+\vspace{-1.5em}
 \label{fig:speedup}
@@
 process, indicating that we can achieve good speedups even when resources are constrained.
+
+While our results pertain to the applicability of Rolling Prefetch to neuroimaging, similar speedups can be expected from other kinds of sequential processing, such as standard text file processing like word count.
 

### 2021-10-05 paper/paper.tex
cut some words

@@
 \subsection{Experiments}
-To evaluate the performance Rolling Prefetch, we conducted 4 experiments: 1) varying
+To evaluate the performance Rolling Prefetch, we conducted four experiments: 1) varying
 the number of files; 2) varying the blocksize; 3) parallel processing of files;  and 4) neuroimaging use-cases.
@@
 
-We chose not to modify the pipeline, in order to determine what is the speedup provided
+We chose not to modify the pipeline in order to determine what is the speedup provided
 by Rolling Prefetch without additional optimization, and in order to be able to speculate on the possible benefits of adapting existing pipelines
@@
 At a very large number of blocks, the performance of both \sfs and Rolling Prefetch is degraded due to
-increased impacts of both S3 storage latency and additionally, in the case of Rolling Prefetch, local storage latency (Figure~\ref{fig:blocksize}).
+increased impacts of both S3 storage latency and, in the case of Rolling Prefetch, local storage latency (Figure~\ref{fig:blocksize}).
 \sfs starts to surpass the speed of Rolling Prefetch, although not substantially (approx.
@@
 
-Conversely, when decreasing the number of blocks we start to see an increase in performance in both \sfs and Rolling Prefetch. This
-is because both benefit from decreased latency to access data on S3, by generally making fewer read calls to it.
+Conversely, when decreasing the number of blocks we see an increase in performance in both \sfs and Rolling Prefetch. This
+is because both benefit from decreased latency to access data on S3, by making fewer read calls to it.
 Rolling Prefetch is found to be faster than \sfs with a peak speedup of 1.24~X at 187 blocks (i.e.
@@
 Currently,
-files are written to specific storage devices based on available space and storage priority in the list.
+files are written to specific storage devices based on available space and storage priority.
 
-With just a simple implementation performed on a computation based on reading, we have observed up to a 1.8$\times$ speedup and a maximum overhead of 1.03$\times$. These observed speedups were
+With just a simple data loading workload, we have observed up to a 1.8$\times$ speedup and a maximum overhead of 1.03$\times$. These observed speedups were
 obtained when we set the cache location to a \texttt{tmpfs} device and may naturally decrease with a slower device such as an SSD.
@@
 implementation to be efficient within parallel processing frameworks, we would need to decouple it from
-the task itself and allow it to communicate with the task scheduler to determine which files to prefetch and
-where.
+the task itself and allow it to communicate with the task scheduler to determine which files to prefetch.
 
